Manuscript Working Draft

### 2023-04-20 manuscript.qmd
cli commit 4

---
title: Manuscript Working Draft

format: 
  pdf: 
    geometry: 
      - top=20mm
      - left=20mm
  docx: default
prefer-html: true
#custom-reference-doc.docx

# documentclass: scrreprt
#     include-in-header:
#       text: |
#         \usepackage{makeidx}
#         \makeindex
#     include-after-body:
#       text: |
#         \printindex
---

@@

[section still in progress...]

## The ABCD Study® Data
Participants enrolled in the ABCD Study include a large cohort of youth (n=11880) aged 9-10 years at baseline and their parents/guardians. The study sample was recruited from household populations in defined catchment areas for each of the 21 study sites across the United States (information regarding funding agencies, recruitment sites, investigators, and project organization can be obtained at the ABCD Study website). The ABCD Study is collecting longitudinal data on a rich variety of outcomes that will enable the construction of realistically-complex   etiological models by incorporating factors from many domains simultaneously. Each new wave of data collection provides the building blocks for conducting probing longitudinal analyses that allow us to characterize normative development, identify variables that presage deviations from prototypic development, and assess a range of outcomes associated with variables of interest. This data includes a neurocognitive battery [@luciana2018a; @thompson2019], mental and physical health assessments [@barch2018], measures of culture and environment [@zucker2018], substance use [@xxxxx], biospecimens [@uban2018], structural and functional brain imaging [@casey2018; @hagler2019], geolocation-based environmental exposure data, wearables, and mobile technology [@bagot2018], and whole genome genotyping [@loughnan2020]. Many of these measures are collected at in-person annual visits, with brain imaging collected at baseline and every other year going forward. A limited number of assessments are collected in semi-annual telephone interviews between in-person visits. Data are publicly released on an annual basis through the NIMH Data Archive. By necessity, the study’s earliest data releases were cross-sectional (i.e., the baseline data), however, the most recent public data release (NDA Release 4.0) contains data collected across three annual assessments, including two imaging assessments (baseline and year 2 follow-up visits).

## Organization of current manuscript
The rich longitudinal nature of the ABCD Study dataset will allow researchers to perform analyses of both methodological and substantive interest. This report describes methods for longitudinal analyses of ABCD Study data that can address its fundamental scientific aims, as well as challenges inherent in a large population-based long-term study of adolescents. The manuscript is organized as follows: xxxxxxxxx.

@@

**Vulnerable periods**. Development normatively progresses from less mature to more mature levels of functioning. However, unique epochs and experiences can alter the course of this idealized form of development. Consider research that shows cannabis use during adolescence is associated with later psychosis to a greater degree than cannabis use initiated later in development [@xxxx]; or, similarly, experimental research on rodents that shows rodent brains to be especially sensitive to the neurotoxic effects of alcohol on brain structure and learning early in development (corresponding to early adolescence in humans)[@xxxx]. These examples highlight the importance of considering the role of vulnerable periods – temporal windows of rapid brain development or remodeling during which the effects of environmental stimuli (e.g. cannabis exposure) on the developing brain may be particularly pronounced– when trying to establish an accurate understanding of the association between exposures and outcomes.

@@

There is growing recognition that statistical models commonly applied to longitudinal data often fail to comport with the developmental theory they are being used to assess (e.g., Curran, Lee, Howard, Lane, & MacCallum, 2012; Hoffman, 2015; Littlefield et al., 2021; xxxxx et al. (xxxx). Specifically, developmental studies typically involve the use of prospective data to inform theories that are concerned with clear within-person (i.e., intraindividual) processes (e.g., how phenotypes change or remain stable within individuals over time) [e.g., see @curran2011]. Despite this, methods generally unsuited for disaggregating between- and within-person effects (e.g., cross-lagged panel models [CLPM]) remain common within various extant literatures. As a result, experts increasingly caution about the need to xxxxxxxx [@xxxxx]. Fortunately, there exists a range of models that have been proposed to tease apart between- and within-person sources of variance across time [see @littlefield20210603; @orth2021]. Most of these contemporary alternatives incorporate time-specific latent variables to capture between-person sources of variance and model within-person deviations around an individual’s mean (or trait) level across time [e.g., RI-CLPM, @hamaker2015; LCM-SR, @curran2014a]. It is important to note however that these models require multiple assessments waves (e.g., four or more to fully specify the LCM-SR), additional expertise to overcome issues with model convergence, and appreciation of modeling assumptions when attempting to adjudicate among potential models in each research context [see @littlefield20210603, for further discussion].

>>>>

title: Manuscript Working Draft

format: 
  pdf: 
    geometry: 
      - top=20mm
      - left=20mm
  docx: default
prefer-html: true

@@

[section still be developed...]

## The ABCD Study® Data
Participants enrolled in the ABCD Study include a large cohort of youth (n=11880) aged 9-10 years at baseline and their parents/guardians. The study sample was recruited from household populations in defined catchment areas for each of the 21 study sites across the United States (information regarding funding agencies, recruitment sites, investigators, and project organization can be obtained at the ABCD Study website). The ABCD Study is collecting longitudinal data on a rich variety of outcomes that will enable the construction of realistically-complex   etiological models by incorporating factors from many domains simultaneously. Each new wave of data collection provides the building blocks for conducting probing longitudinal analyses that allow us to characterize normative development, identify variables that presage deviations from prototypic development, and assess a range of outcomes associated with variables of interest. This data includes a neurocognitive battery [@luciana2018a; @thompson2019], mental and physical health assessments [@barch2018], measures of culture and environment [@zucker2018], substance use [add citation], biospecimens [@uban2018], structural and functional brain imaging [@casey2018; @hagler2019], geolocation-based environmental exposure data, wearables, and mobile technology [@bagot2018], and whole genome genotyping [@loughnan2020]. Many of these measures are collected at in-person annual visits, with brain imaging collected at baseline and every other year going forward. A limited number of assessments are collected in semi-annual telephone interviews between in-person visits. Data are publicly released on an annual basis through the NIMH Data Archive. By necessity, the study’s earliest data releases were cross-sectional (i.e., the baseline data), however, the most recent public data release (NDA Release 4.0) contains data collected across three annual assessments, including two imaging assessments (baseline and year 2 follow-up visits).

## Organization of current manuscript
The rich longitudinal nature of the ABCD Study dataset will allow researchers to perform analyses of both methodological and substantive interest. This report describes methods for longitudinal analyses of ABCD Study data that can address its fundamental scientific aims, as well as challenges inherent in a large population-based long-term study of adolescents. The manuscript is organized as follows:

[section still be developed...]

@@

**Vulnerable periods**. Development normatively progresses from less mature to more mature levels of functioning. However, unique epochs and experiences can alter the course of this idealized form of development. Consider research that shows cannabis use during adolescence is associated with later psychosis to a greater degree than cannabis use initiated later in development [add citation]; or, similarly, experimental research on rodents that shows rodent brains to be especially sensitive to the neurotoxic effects of alcohol on brain structure and learning early in development (corresponding to early adolescence in humans)[add citation]. These examples highlight the importance of considering the role of vulnerable periods – temporal windows of rapid brain development or remodeling during which the effects of environmental stimuli (e.g. cannabis exposure) on the developing brain may be particularly pronounced– when trying to establish an accurate understanding of the association between exposures and outcomes.

@@

There is growing recognition that statistical models commonly applied to longitudinal data often fail to comport with the developmental theory they are being used to assess (e.g., Curran, Lee, Howard, Lane, & MacCallum, 2012; Hoffman, 2015; Littlefield et al., 2021. Specifically, developmental studies typically involve the use of prospective data to inform theories that are concerned with clear within-person (i.e., intraindividual) processes (e.g., how phenotypes change or remain stable within individuals over time) [e.g., see @curran2011]. Despite this, methods generally unsuited for disaggregating between- and within-person effects (e.g., cross-lagged panel models [CLPM]) remain common within various extant literatures. As a result, experts increasingly caution about the need to xxxxxxxx [add citation]. Fortunately, there exists a range of models that have been proposed to tease apart between- and within-person sources of variance across time [see @littlefield20210603; @orth2021]. Most of these contemporary alternatives incorporate time-specific latent variables to capture between-person sources of variance and model within-person deviations around an individual’s mean (or trait) level across time [e.g., RI-CLPM, @hamaker2015; LCM-SR, @curran2014a]. It is important to note however that these models require multiple assessments waves (e.g., four or more to fully specify the LCM-SR), additional expertise to overcome issues with model convergence, and appreciation of modeling assumptions when attempting to adjudicate among potential models in each research context [see @littlefield20210603, for further discussion].
